---
title: "The Costs of Michigan Healthcare Consolidation"
author: "Laine Cibulskis"
date: "6-6-2026"
format:
  html:
    theme: cosmo
---

Plan for second memo:
- Narrow pitch: focus on connecting hospital finances to patients
- Contextualize every table with context on why it matters:
- what the data mean
- why the frames were chosen (i.e. why Illinois)
- rename columns to be easily understandable
- merging some tables may make it easier to see, plus hiding some extra dense code:
- like merging the two hospital analyses into one chart
- why do rising/high hospital expenditures affect Michiganders?
- add some additional data analysis about insurance disparities and patient outcomes/care

Intersting additional variables:
Row 2499 preventable hospital stays rate:
"Hospitalization for ambulatory care-sensitive conditions, diagnoses usually treatable in outpatient services, suggests that the quality of care provided in the outpatient setting was not accessible. The measure may also represent a tendency to overuse emergency room and urgent care as a main source of care."
medcr_ffs_prev_hosp_rate_22
medcr_ffs_prev_hosp_rate_21
hosp_beds_23
hosp_beds_22
stgh_hosp_beds_23
stgh_hosp_beds_22
stngh_hosp_beds_23
stngh_hosp_beds_22
lth_hosp_beds_23
lth_hosp_beds_22

## Hypothesis/possible headline
Patients pay the price as Michigan healthcare systems consolidate

## In two sentences, what is this story about? 
Michigan hospital systems are getting bigger, but have patients paid a price for consolidation? This story will explore the shrinking number of players in Michigan healthcare in conjunction with data on the business of healthcare: hospital expenditures, insurance reimbursements and patient costs.

## In one sentence, why tell this story NOW?
300,000 residents [nearly](https://www.detroitnews.com/story/business/2026/05/27/blue-cross-michigan-medicine-reach-deal/90274740007/) lost their access to the state's premier health system thanks to an insurance dispute [in June](https://www.detroitnews.com/story/business/2026/05/06/blue-cross-tells-250000-michigan-medicine-patients-find-new-doctors/89960599007/), healthcare costs [continue to soar](https://pmc.ncbi.nlm.nih.gov/articles/PMC12920308/) and a $2.2 billion hospital construction project just reached [20 stories tall last month](https://www.henryford.com/news/2026/05/henry-ford-hospital-construction-milestone#): it's time explain the juxtaposition between shiny buildings and growing systems with patient struggles.

## Six things (5W and H) you learned about this story 
Who - hospitals, insurance companies and patients

What - mergers making systems larger, hospital expenditures growing

When - decades of growing costs and consolidation 

Where - Michigan and its counties, plus comparisons to the U.S.

Why - Michigan has high healthcare spending compared to the nation, but Michiganders still need to know how consolidation has actually affected their finances and health, beyond what the pressers from healthcare companies and daily news can tell them. Additionally, the Detroit News has identified a gap in its data coverage of healthcare, in addition to a gap in its business of healthcare coverage, which is usually covered on an as-needed basis

## Key data analysis points (to be updated as training continues)
Michigan is ranked No. 14 in the nation, excluding CT (to be addressed later), for per hospital short term general hospital expenditures by state.

$342,074,250 is Michigan's per hospital value--exactly the third quartile. Meaning 3/4 of the U.S. states (again, minus CT), have lower per hospital expenditures for short term general hospitals.

**Each** short term general hospital in Michigan spends about $50 million more than the nation's average. 

## Three people I can interview for this story are
**Thomas Buchmueller**, University of Michigan health economist

**Bill Farmer**, healthcare advocate who co-founded the Michigan For Single Payer Healthcare grassroots organization with **Eli Rubin**

**Carl St. Amour**, Michigan Medicine's Director of Revenue, Reporting, and Payment Strategies

**Mathhew Cox**, Chief Financial Officer at Corewell Health

Founders of Michigan's New Day Foundation for Families **Michael Spehn and Gina Kell Spehn** may be able to connect me with patients receiving high-cost specialty care. New Day focuses on working with hospitals to reach patients nearing financial breaking points, especially when it comes to cancer treatment. Many board members have had personal experiences with cancer and costs as well

## Data you will be using to inform the story: 
Trends in hospital expenditures and other healthcare data from:

[Health Resources and Services Administration Area Health Resource Files](https://data.hrsa.gov/data/download)

[Wayne County healthcare consolidation data](https://detroitdata.org/dataset/market-saturation-utilization)

[Hospital-level expenditure data](https://tool.nashp.org/)

## What are the visual possibilities?
County-by county data from the Area Health Resource Files could be used to create a gradient map with a dropdown menu to show different variables for each Michigan county. There could be a scroller for change over time in different metrics such as hospital expenditures and patient outcomes.

A locator map could show detailed information about Detroit healthcare consolidation (imagine a map with points color coded by offices owned by differnet entities); another locator map could also locations of hospitals throughout the state with key expenditure points.

A collaboration with a photographer could result in a patient photo story to add a human element.

## How did you find this story? Summarize any previous coverage. 
Last week, Blue Cross Blue Shield of Michigan finally reached a settlement with the University of Michigan Health (Michigan Medicine). The contract renegotiation between the two systems, which allows patients under the insurer to receive care at Michigan Medicine, had nearly failed, threatening 300,000 residents with a loss of care from one of the state's most reputable health systems. As in many industries, consolidation has taken over healthcare. A 2022 Michigan health care merger created a combined 22-hospital system, Corewell Health, with claims of improving patient care. In 2023, Michigan Medicine acquired another health system as well. Still, insurance companies blame hospitals for rising costs, and insurance companies blame hospitals. This April, Michigan Medicine sang praise for the results of the 2023 merger. 

But a [May KFF](https://www.kff.org/health-costs/hospital-prices-have-risen-much-faster-for-private-insurance-than-medicare-since-2019/) study showed hospital prices are rising much faster for private insurance than Medicare since 2019. Another [KFF briefing from 2024](https://www.kff.org/health-costs/ten-things-to-know-about-consolidation-in-health-care-provider-markets/) explains that a large body of evidence point to healthcare conslidiation increasing costs, but that the verdict on patient outcomes is less straightforward.

Michigan Medicine [claims](https://prod.crainsdetroit.com/health-care/whats-behind-blue-cross-michigan-medicine-contract-dispute) it should be rewarded with better contracts thanks to its quality of care and focus on specialty medicine. But how has consolidation really worked out for patients in Michigan, from their health to their pocketbooks? Especially as Detroit's hospital buildings get shinier at taller, as with Henry Ford Hospital's in-progress $2.2 billion, 20-story [expansion]((https://www.henryford.com/news/2026/05/henry-ford-hospital-construction-milestone#)?

## What other information do you need to gather? 
* Academic input on the nuances of healthcare market consolidation as it relates specifically to Michigan
* Corporate and patient input
* Insurance input (private, Medicare/Medicaid)
* Continued exploration of data

## Who do you need to contact? 
Kim North Shine, senior public relations representative at the University of Michigan who specializes in public health as one of her topic areas, may be able to help me get in touch with both academic and healthcare sources. Connecting with the advocacy groups mentioned above will also help kick start the reporting side of the story.

# 5/31: The Idea

Last week, Blue Cross Blue Shield of Michigan finally reached a settlement with the University of Michigan Health (Michigan Medicine). The contract renegotiation between the two systems, which allows patients under the insurer to receive care at Michigan Medicine, had nearly failed, threatening 300,000 residents with a loss of care from one of the state's most reputable health systems. As in many industries, consolidation has taken over healthcare. A 2022 Michigan health care merger created a combined 22-hospital system, Corewell Health, with claims of improving patient care. In 2023, Michigan Medicine acquired another health system as well. Still, insurance companies blame hospitals for rising costs, and insurance companies blame hospitals. This April, Michigan Medicine sang praise for the results of the 2023 merger. 

But a May KFF study showed hospital prices are rising much faster for private insurance than Medicare since 2019. Michigan Medicine claims it should be rewarded with better contracts thanks to its quality of care and focus on specialty medicine. But how has consolidation really worked out for patients, from their health to their pocketbooks? Especially as Detroit's hospital buildings get shinier (see Henry Ford Hospital's recent $2.2 billion expansion). 

Enter: Health Resources and Services Administration data. The HRSA's Area Health Resource files provide a wealth of information on hospitals by county, state and for the entire nation. An initial look reveals that in just one year, hospital expenditures per non-general and long term hospital in Washtenaw County (includes Ann Arbor, where Michigan Health is based) grew nearly $30 million to reach over $1 billion per hospital as of the most recent data available from 2023 (we have a data lag here according to the documentation). Why is that so much larger than Illinois' largest county for the same measure, at about $600 million? How quickly have these expenditures risen, county-by-county? Is Michigan rising faster than other states, and how do increases show in areas with newly-merged hospitals? Have the mergers good for expenditures? And patients, such as with care quality metrics also in the data?

# Data analysis: hospital expenditures in 2025

# Setup

## Packages

```{r}
#install.packages("tidyverse")
#install.packages("janitor")
#install.packages("rio")
#install.packages("googledrive")
#install.packages("readxl")
```

## Libraries

```{r}
#| include: false
library(tidyverse)
library(janitor)
library(rio)
library(googledrive)
library(readxl)
library(dplyr)
```

## Read the files

```{r}
AHFR2025 <- read.csv("C:/Users/laine/OneDrive/Desktop/Code/DJNF/laine-cibulskis/data/AHRF2025.csv")
AHFR2025_MI <- read.csv("C:/Users/laine/OneDrive/Desktop/Code/DJNF/laine-cibulskis/data/AHRF2025_MI.csv")
AHFR25_IL <- read.csv("C:/Users/laine/OneDrive/Desktop/Code/DJNF/laine-cibulskis/data/AHRF2025_IL.csv")
```

## Documentation

```{r}
AHRF2025documentation <- read_excel("C:/Users/laine/OneDrive/Desktop/Code/DJNF/laine-cibulskis/documentation/AHRF_USER_TECH_2024-2025/Technical Documentation/AHRF 2024-2025 Technical Documentation.xlsx")
# Expenditures
AHRF2025documentation[c(3313, 3314, 3315, 3316, 3319, 3320, 3321, 3322), ]

```
# County-level Michigan analysis

### 2025 AHRF data: average expenditures per short-term general (generalhosp) hospital in Michigan counties

```{r}
generalhosp23_MI <- AHRF2025_MI |>
  select(cnty_name,stgh_reprtd_fac_exp_thou_23, stgh_rptng_exp_23) |>
  rename(County = cnty_name, Hospitals = stgh_rptng_exp_23, Expenditures = stgh_reprtd_fac_exp_thou_23) |>
  mutate("Expenditures per Hospital" = ((Expenditures * 1000) / Hospitals)) |>
  mutate_all(~replace(., is.nan(.), 0)) |>
  arrange(desc("Expenditures per Hospital")
)
```

### 2025 AHRF data: average expenditures per short-term non-general and long term (extracare) hospitals in Michigan counties
*The AHRF 2024-2025 User Guide explains that short term non-general hospitals provide specialized care, and the majority of their patients stay for fewer than 30 days. Long term hospitals may provide either non specialized or specialized care, and the majority of their patients stay for 30 or more days.*

```{r}
extracare23_MI <- AHRF2025_MI |>
  select(cnty_name,stnglth_reprtd_fac_exp_thou_23, stnglth_rptng_exp_23) |>
  rename(County = cnty_name, Hospitals = stnglth_rptng_exp_23, Expenditures = stnglth_reprtd_fac_exp_thou_23) |>
  mutate("Expenditures per Hospital" = ((Expenditures * 1000) / Hospitals)) |>
  mutate_all(~replace(., is.nan(.), 0)) |>
  arrange(desc("Expenditures per Hospital")
)
```

## Quick IL comparison

### 2025 AHRF data: average expenditures per short-term general (generalhosp) hospital in Illinois counties

```{r}
generalhosp23_IL <- AHRF2025_IL |>
  select(cnty_name,stgh_reprtd_fac_exp_thou_23, stgh_rptng_exp_23) |>
  rename(County = cnty_name, Hospitals = stgh_rptng_exp_23, Expenditures = stgh_reprtd_fac_exp_thou_23) |>
  mutate("Expenditures per Hospital" = ((Expenditures * 1000) / Hospitals)) |>
  mutate_all(~replace(., is.nan(.), 0)) |>
  arrange(desc("Expenditures per Hospital")
)
```

### 2025 AHRF data: average expenditures per short-term non-general and long term (extracare) hospitals in Michigan counties

```{r}
extracare23_IL <- AHRF2025_IL |>
  select(cnty_name,stnglth_reprtd_fac_exp_thou_23, stnglth_rptng_exp_23) |>
  rename(County = cnty_name, Hospitals = stnglth_rptng_exp_23, Expenditures = stnglth_reprtd_fac_exp_thou_23) |>
  mutate("Expenditures per Hospital" = ((Expenditures * 1000) / Hospitals)) |>
  mutate_all(~replace(., is.nan(.), 0)) |>
  arrange(desc("Expenditures per Hospital")
)
```

# National comparison by state

## Short-term general hospitals

### Total short-term general hospital expenditures by state
```{r}
stgh_23_exp_by_state <- 
  AHRF2025 |>
  select(st_name_abbrev, stgh_reprtd_fac_exp_thou_23) |>
  group_by(st_name_abbrev) |>
  summarise(stgh_23_by_state = (sum((stgh_reprtd_fac_exp_thou_23 * 1000)))
  )
```

### Number of short-term general hospitals by state

```{r}
stgh_23_by_state <- 
  AHRF2025 |>
  select(st_name_abbrev, stgh_rptng_exp_23) |>
  group_by(st_name_abbrev) |>
  summarise(stgh_23_by_state = (sum(stgh_rptng_exp_23))) |>
  arrange(desc(stgh_23_by_state)
  )
```

### Joining the two new data frames

```{r}
stgh_23 <-
merge(x = stgh_23_exp_by_state, y = stgh_23_by_state, by = "st_name_abbrev", .keep_all=TRUE)
```

### Total short-term general hospital expenditures by state, per hospital in the state

```{r}
stgh_23 |>
  mutate(stgh_23_exp_by_state_ph = (stgh_23_by_state.x / stgh_23_by_state.y)) |>
  arrange(desc(stgh_23_exp_by_state_ph)
  )
```

### Save total short term general hospital expenditures by state, per hospital in the state to use in other operations

```{r}
# Renaming
stgh_23 <-
  stgh_23 |>
    rename(
    Expenditures = stgh_23_by_state.x,
    Hospitals = stgh_23_by_state.y,
    State = st_name_abbrev
    )

```

```{r}
generalhosp23 <-
  stgh_23 |>
  select(State, Hospitals, Expenditures) |>
  mutate("Expenditures per Hospital" = Expenditures / Hospitals) |>
  arrange(desc("Expenditures per Hospital"))
```

# National comparison by state

## Short-term non-general and long-term hospitals

### Total short-term non-general and long-term hospital expenditures by state
```{r}
stnglth_23_exp_by_state <- 
  AHRF2025 |>
  select(st_name_abbrev, stnglth_reprtd_fac_exp_thou_23) |>
  group_by(st_name_abbrev) |>
  summarise(stnglth_23_by_state = (sum((stnglth_reprtd_fac_exp_thou_23 * 1000)))
  )
```

### Number of short-term non-general and long-term hospitals by state

```{r}
stnglth_23_by_state <- 
  AHRF2025 |>
  select(st_name_abbrev, stnglth_rptng_exp_23) |>
  group_by(st_name_abbrev) |>
  summarise(stnglth_23_by_state = (sum(stnglth_rptng_exp_23))) |>
  arrange(desc(stnglth_23_by_state)
  )
```

### Joining the two new data frames

```{r}
stnglth_23 <-
merge(x = stnglth_23_exp_by_state, y = stnglth_23_by_state, by = "st_name_abbrev", .keep_all=TRUE)
```

### Total short-term general hospital expenditures by state, per hospital in the state

```{r}
stnglth_23 |>
  mutate(stnglth_23_exp_by_state_ph = (stnglth_23_by_state.x / stnglth_23_by_state.y)) |>
  arrange(desc(stnglth_23_exp_by_state_ph)
  )
```

### Save total short term general hospital expenditures by state, per hospital in the state to use in other operations

```{r}
# Renaming
stnglth_23 <-
  stnglth_23 |>
    rename(
    Expenditures = stnglth_23_by_state.x,
    Hospitals = stnglth_23_by_state.y,
    State = st_name_abbrev
    )

```

```{r}
extracare23 <-
  stnglth_23 |>
  select(State, Hospitals, Expenditures) |>
  mutate("Expenditures per Hospital" = Expenditures / Hospitals) |>
  arrange(desc("Expenditures per Hospital"))
```


```{r}
stnglth23_exp_by_state <- 
  AHRF2025 |>
  select(st_name_abbrev, stnglth_reprtd_fac_exp_thou_23) |>
  group_by(st_name_abbrev) |>
  mutate(stnglth_23_by_state = (sum((stnglth_reprtd_fac_exp_thou_23 * 1000)))) |>
  mutate(stnglth_23_by_state = (sum(stnglth_rptng_exp_23)
  )
```

### Create short term general hospitals by state

```{r}
stnglth_23_by_state <- 
  AHRF2025 |>
  select(st_name_abbrev, stnglth_rptng_exp_23) |>
  group_by(st_name_abbrev) |>
  
  )
```

### Joining the two new data frames

```{r}
stnglth23 <-
  stnglth_23_by_state |>
  select(st_name_abbrev, stnglth_rptng_exp_23, stnglth_23_by_state)
merge(x = stnglth_rptng_exp_23, y = stnglth_23_by_state, by = "st_name_abbrev", .keep_all=TRUE)
```

### Total short term general hospital expenditures by state, per hospital in the state (don't need to run again because data frame was saved with new headers)

```{r}
stnglth_23 |>
  mutate(stnglth_23_exp_by_state_ph = (stnglth_23_by_state.x / stnglth_23_by_state.y)) |>
  arrange(desc(stnglth_23_exp_by_state_ph)
  )
```

### Save total short term general hospital expenditures by state, per hospital in the state to use in other operations

```{r}
# Renaming
stnglth_23 <-
  stnglth_23 |>
    rename(
    Expenditures = stnglth_23_by_state.x,
    Hospitals = stnglth_23_by_state.y,
    State = st_name_abbrev
    )

```

```{r}
extracare23 <-
  stgh_23 |>
  select(State, Hospitals, Expenditures) |>
  mutate("Expenditures per Hospital" = Expenditures / Hospitals) |>
  arrange(desc("Expenditures per Hospital"))
```

### General summary

```{r}
summary(shortterm23$"Expenditures per Hospital")
```

### Michigan is ranked No. 14 in the nation, excluding CT (to be addressed later), for per hospital short term general hospital expenditures by state.

### $342,074,250 is Michigan's per hospital value--exactly the third quartile. Meaning 3/4 of the U.S. states (again, minus CT), have lower per hospital expenditures for short term general hospitals.

### Michigan difference from mean calculation

```{r}
342074250 - 291256410
```

## Short-term non-general and long-term hospitals



# Creating Charts

## Michigan by county

### Join Michigan data by county for map

```{r}
full_join(
  x, stgh_exp_ph
  y, 
  by = NULL,
  copy = FALSE,
  suffix = c(".x", ".y"),
  ...,
  keep = NULL,
  na_matches = c("na", "never"),
  multiple = "all",
  relationship = NULL
)
```

### Load libraries and API key

```{r}
library(DatawRappr)
library(tidyverse)
datawrapper_auth(api_key =  "qWGtNy2YSjg6j2gfzgCSQ3iYFeeciWYTJTuWyVfTfWqWI7jbNHs9HOcmYIOsc6X2", overwrite=TRUE)
```

### Call in data and display head

my_chart <- dw_create_chart()

  rename(
    State = st_name_abbrev
    Revenue, short term general hospitals = stgh_23_by_state.x
    Total short term general hospitals = stgh_23_by_state.y