---
title: "The Costs of Michigan Healthcare Consolidation"
author: "Laine Cibulskis"
date: "6-6-2026"
format:
  html:
    theme: cosmo
---
# 6/3: The Memo, Second Draft

Plan for second memo:
- Narrow pitch: focus on connecting hospital finances to patients
- Contextualize every table with context on why it matters:
- what the data mean
- why the frames were chosen (i.e. why Illinois)
- rename columns to be easily understandable
- merging some tables may make it easier to see, plus hiding some extra dense code:
- like merging the two hospital analyses into one chart
- why do rising/high hospital expenditures affect Michiganders?
- add some additional data analysis about insurance disparities and patient outcomes/care

Intersting additional variables:
Row 2499 preventable hospital stays rate:
"Hospitalization for ambulatory care-sensitive conditions, diagnoses usually treatable in outpatient services, suggests that the quality of care provided in the outpatient setting was not accessible. The measure may also represent a tendency to overuse emergency room and urgent care as a main source of care."
medcr_ffs_prev_hosp_rate_22
medcr_ffs_prev_hosp_rate_21
hosp_beds_23
hosp_beds_22
stgh_hosp_beds_23
stgh_hosp_beds_22
stngh_hosp_beds_23
stngh_hosp_beds_22
lth_hosp_beds_23
lth_hosp_beds_22

## Datawrapper Embed test

<iframe title="Unemployment trends across the U.S. in 2024." aria-label="Choropleth map" id="datawrapper-chart-gmHRV" src="https://datawrapper.dwcdn.net/gmHRV/1/" scrolling="no" frameborder="0" style="width: 0; min-width: 100% !important; border: none;" height="" data-external="1"></iframe><script type="text/javascript">(function(){function e(){window.addEventListener(`message`,function(e){if(e.data[`datawrapper-height`]!==void 0){var t=document.querySelectorAll(`iframe`);for(var n in e.data[`datawrapper-height`])for(var r=0,i;i=t[r];r++)if(i.contentWindow===e.source){var a=e.data[`datawrapper-height`][n]+`px`;i.style.height=a}}})}e()})();</script>

# 6/1: The Memo, First Draft

## Slug
The Costs of Michigan Healthcare Consolidation

## Hypothesis/possible headline
Patients pay the price as Michigan healthcare systems consolidate

## In two sentences, what is this story about? 
Michigan hospital systems are getting bigger, but have patients paid a price for consolidation? This story will explore the shrinking number of players in Michigan healthcare in conjunction with data on the business of healthcare: hospital expenditures, insurance reimbursements and patient costs.

## In one sentence, why tell this story NOW?
300,000 residents [nearly](https://www.detroitnews.com/story/business/2026/05/27/blue-cross-michigan-medicine-reach-deal/90274740007/) lost their access to the state's premier health system thanks to an insurance dispute [in June](https://www.detroitnews.com/story/business/2026/05/06/blue-cross-tells-250000-michigan-medicine-patients-find-new-doctors/89960599007/), healthcare costs [continue to soar](https://pmc.ncbi.nlm.nih.gov/articles/PMC12920308/) and a $2.2 billion hospital construction project just reached [20 stories tall last month](https://www.henryford.com/news/2026/05/henry-ford-hospital-construction-milestone#): it's time explain the juxtaposition between shiny buildings and growing systems with patient struggles.

## Six things (5W and H) you learned about this story 
Who - hospitals, insurance companies and patients

What - mergers making systems larger, hospital expenditures growing

When - decades of growing costs and consolidation 

Where - Michigan and its counties, plus comparisons to the U.S.

Why - Michigan has high healthcare spending compared to the nation, but Michiganders still need to know how consolidation has actually affected their finances and health, beyond what the pressers from healthcare companies and daily news can tell them. Additionally, the Detroit News has identified a gap in its data coverage of healthcare, in addition to a gap in its business of healthcare coverage, which is usually covered on an as-needed basis

## Key data analysis points (to be updated as training continues)
Michigan is ranked No. 14 in the nation, excluding CT (to be addressed later), for per hospital short term general hospital expenditures by state.

$342,074,250 is Michigan's per hospital value--exactly the third quartile. Meaning 3/4 of the U.S. states (again, minus CT), have lower per hospital expenditures for short term general hospitals.

**Each** short term general hospital in Michigan spends about $50 million more than the nation's average. 

## Three people I can interview for this story are
**Thomas Buchmueller**, University of Michigan health economist

**Bill Farmer**, healthcare advocate who co-founded the Michigan For Single Payer Healthcare grassroots organization with **Eli Rubin**

**Carl St. Amour**, Michigan Medicine's Director of Revenue, Reporting, and Payment Strategies

**Mathhew Cox**, Chief Financial Officer at Corewell Health

Founders of Michigan's New Day Foundation for Families **Michael Spehn and Gina Kell Spehn** may be able to connect me with patients receiving high-cost specialty care. New Day focuses on working with hospitals to reach patients nearing financial breaking points, especially when it comes to cancer treatment. Many board members have had personal experiences with cancer and costs as well

## Data you will be using to inform the story: 
Trends in hospital expenditures and other healthcare data from:

[Health Resources and Services Administration Area Health Resource Files](https://data.hrsa.gov/data/download)

[Wayne County healthcare consolidation data](https://detroitdata.org/dataset/market-saturation-utilization)

[Hospital-level expenditure data](https://tool.nashp.org/)

## What are the visual possibilities?
County-by county data from the Area Health Resource Files could be used to create a gradient map with a dropdown menu to show different variables for each Michigan county. There could be a scroller for change over time in different metrics such as hospital expenditures and patient outcomes.

A locator map could show detailed information about Detroit healthcare consolidation (imagine a map with points color coded by offices owned by differnet entities); another locator map could also locations of hospitals throughout the state with key expenditure points.

A collaboration with a photographer could result in a patient photo story to add a human element.

## How did you find this story? Summarize any previous coverage. 
Last week, Blue Cross Blue Shield of Michigan finally reached a settlement with the University of Michigan Health (Michigan Medicine). The contract renegotiation between the two systems, which allows patients under the insurer to receive care at Michigan Medicine, had nearly failed, threatening 300,000 residents with a loss of care from one of the state's most reputable health systems. As in many industries, consolidation has taken over healthcare. A 2022 Michigan health care merger created a combined 22-hospital system, Corewell Health, with claims of improving patient care. In 2023, Michigan Medicine acquired another health system as well. Still, insurance companies blame hospitals for rising costs, and insurance companies blame hospitals. This April, Michigan Medicine sang praise for the results of the 2023 merger. 

But a [May KFF](https://www.kff.org/health-costs/hospital-prices-have-risen-much-faster-for-private-insurance-than-medicare-since-2019/) study showed hospital prices are rising much faster for private insurance than Medicare since 2019. Another [KFF briefing from 2024](https://www.kff.org/health-costs/ten-things-to-know-about-consolidation-in-health-care-provider-markets/) explains that a large body of evidence point to healthcare conslidiation increasing costs, but that the verdict on patient outcomes is less straightforward.

Michigan Medicine [claims](https://prod.crainsdetroit.com/health-care/whats-behind-blue-cross-michigan-medicine-contract-dispute) it should be rewarded with better contracts thanks to its quality of care and focus on specialty medicine. But how has consolidation really worked out for patients in Michigan, from their health to their pocketbooks? Especially as Detroit's hospital buildings get shinier at taller, as with Henry Ford Hospital's in-progress $2.2 billion, 20-story [expansion]((https://www.henryford.com/news/2026/05/henry-ford-hospital-construction-milestone#)?

## What other information do you need to gather? 
* Academic input on the nuances of healthcare market consolidation as it relates specifically to Michigan
* Corporate and patient input
* Insurance input (private, Medicare/Medicaid)
* Continued exploration of data

## Who do you need to contact? 
Kim North Shine, senior public relations representative at the University of Michigan who specializes in public health as one of her topic areas, may be able to help me get in touch with both academic and healthcare sources. Connecting with the advocacy groups mentioned above will also help kick start the reporting side of the story.

# 5/31: The Idea

Last week, Blue Cross Blue Shield of Michigan finally reached a settlement with the University of Michigan Health (Michigan Medicine). The contract renegotiation between the two systems, which allows patients under the insurer to receive care at Michigan Medicine, had nearly failed, threatening 300,000 residents with a loss of care from one of the state's most reputable health systems. As in many industries, consolidation has taken over healthcare. A 2022 Michigan health care merger created a combined 22-hospital system, Corewell Health, with claims of improving patient care. In 2023, Michigan Medicine acquired another health system as well. Still, insurance companies blame hospitals for rising costs, and insurance companies blame hospitals. This April, Michigan Medicine sang praise for the results of the 2023 merger. 

But a May KFF study showed hospital prices are rising much faster for private insurance than Medicare since 2019. Michigan Medicine claims it should be rewarded with better contracts thanks to its quality of care and focus on specialty medicine. But how has consolidation really worked out for patients, from their health to their pocketbooks? Especially as Detroit's hospital buildings get shinier (see Henry Ford Hospital's recent $2.2 billion expansion). 

Enter: Health Resources and Services Administration data. The HRSA's Area Health Resource files provide a wealth of information on hospitals by county, state and for the entire nation. An initial look reveals that in just one year, hospital expenditures per non-general and long term hospital in Washtenaw County (includes Ann Arbor, where Michigan Health is based) grew nearly $30 million to reach over $1 billion per hospital as of the most recent data available from 2023 (we have a data lag here according to the documentation). Why is that so much larger than Illinois' largest county for the same measure, at about $600 million? How quickly have these expenditures risen, county-by-county? Is Michigan rising faster than other states, and how do increases show in areas with newly-merged hospitals? Have the mergers good for expenditures? And patients, such as with care quality metrics also in the data?

# Data analysis for the Detroit News

Examine hospital expenditures over time by county, the effects of consolidation

### Load packages - do this once

```{r}
#install.packages("tidyverse")
#install.packages("janitor")
#install.packages("rio")
#install.packages("googledrive")
#install.packages("readxl")
```

### Load libraries, do this every time

```{r}
#| include: false
library(tidyverse)
library(janitor)
library(rio)
library(googledrive)
library(readxl)
library(dplyr)
```

# Loading data AHRFSN data for 2024 and 2025

```{r}
#| include: false
ahrfsn2025 <- read_csv("https://raw.githubusercontent.com/djnf-data-2026/laine-cibulskis/refs/heads/main/data/ahrfsn2025.csv")

ahrfsn2024 <- read_csv("https://raw.githubusercontent.com/djnf-data-2026/laine-cibulskis/refs/heads/main/data/ahrfsn2024.csv")

#glimpse(ahrfsn2025)

```

# Loading AHRF 2025 and 2024 files

```{r}

AHRF2025_MI <- read_csv("https://raw.githubusercontent.com/djnf-data-2026/laine-cibulskis/refs/heads/main/data/AHRF2025_MI.csv")


AHRF2024_MI <- read_csv("https://raw.githubusercontent.com/djnf-data-2026/laine-cibulskis/refs/heads/main/data/AHRF2024_MI.csv")

```

# Data Wrangling Notes from Rob

You don't need to run this code below -- I kept it to show you how I loaded the AHRF2025 file from a Google Drive link and then created the filtered Michigan files

```{r}
#| eval: false
#| include: false
googledrive::drive_auth() #you may need to authorize your browser

# Here's the file: https://drive.google.com/file/d/1K8DzTY5yTNI0EeIUpadeCrE552tcxRsj/view?usp=sharing

drive_download(
  as_id("1K8DzTY5yTNI0EeIUpadeCrE552tcxRsj"),  # copied from the file link
  path = "../data/AHRF2025.csv", #a random data path - you can select any folder you want
  overwrite = TRUE
)
AHRF2025 <- read_csv("../data/AHRF2025.csv")|> 
  clean_names()

AHRF2025_MI <- AHRF2025 |> 
  filter(st_name_abbrev == "MI")

#write_csv(AHRF2025_MI, "../data/AHRF2025_MI.csv" )

```

# Load AHRF2024 file

Same thing as above

```{r}
#| eval: false
#| include: false
#https://drive.google.com/file/d/1GgVgzo7DSv1pn8BjCuhygmILEjGT04Ja/view?usp=sharing
#googledrive::drive_auth()
#drive_download(
  #as_id("1GgVgzo7DSv1pn8BjCuhygmILEjGT04Ja"),  # from the share link
  #path = "./data/AHRF2024.csv",
  #overwrite = TRUE
#)
#AHRF2024 <- read_csv("./data/AHRF2024.csv")|> 
  #clean_names()

#AHRF2024_MI <- AHRF2024 |> 
  #filter(st_name_abbrev == "MI")

#write_csv(AHRF2024_MI, "./data/AHRF2024_MI.csv" )

```

# 5/31: Inital Analysis for Memo

## Create Illinois files for comparison, 2025

```{r}
#| eval: false
#| include: false
#library(googledrive)
#googledrive::drive_auth() #you may need to authorize your browser

# Here's the file: https://drive.google.com/file/d/1K8DzTY5yTNI0EeIUpadeCrE552tcxRsj/view?usp=sharing
#googledrive::drive_auth()
#drive_download(
  #as_id("1K8DzTY5yTNI0EeIUpadeCrE552tcxRsj"),  # copied from the file link
  #path = "C:/Users/laine/OneDrive/Desktop/Code/DJNF/laine-cibulskis/data/AHRF2025.csv", #a random data path - you can select any folder you want
  #overwrite = TRUE
)
#AHRF2025 <- read_csv("C:/Users/laine/OneDrive/Desktop/Code/DJNF/laine-cibulskis/data/AHRF2025.csv")|> 
  #clean_names()

#AHRF2025_IL <- AHRF2025 |> 
  #filter(st_name_abbrev == "IL")

#write_csv(AHRF2025_IL, "C:/Users/laine/OneDrive/Desktop/Code/DJNF/laine-cibulskis/data/AHRF2025_IL.csv" )

```

## Create Illinois files for comparison, 2024

```{r}
#| eval: false
#| include: false
#https://drive.google.com/file/d/1GgVgzo7DSv1pn8BjCuhygmILEjGT04Ja/view?usp=sharing
#googledrive::drive_auth()
#drive_download(
  #as_id("1GgVgzo7DSv1pn8BjCuhygmILEjGT04Ja"),  # from the share link
  #path = "C:/Users/laine/OneDrive/Desktop/Code/DJNF/laine-cibulskis/data/AHRF2024.csv",
  #overwrite = TRUE
#)
#AHRF2024 <- read_csv("C:/Users/laine/OneDrive/Desktop/Code/DJNF/laine-cibulskis/data/AHRF2024.csv")|> 
  clean_names()

#AHRF2024_IL <- AHRF2024 |> 
  filter(st_name_abbrev == "IL")

#write_csv(AHRF2024_IL, "C:/Users/laine/OneDrive/Desktop/Code/DJNF/laine-cibulskis/data/AHRF2024_IL.csv" )


```

## Load AHRF files for Illinois and all other counties in the nation

```{r}
AHRF2024_IL <- read_csv("https://raw.githubusercontent.com/djnf-data-2026/laine-cibulskis/refs/heads/main/data/AHRF2024_IL.csv")

AHRF2025_IL <- read_csv("https://raw.githubusercontent.com/djnf-data-2026/laine-cibulskis/refs/heads/main/data/AHRF2025_IL.csv")

AHRF2024 <- read_csv("https://raw.githubusercontent.com/djnf-data-2026/laine-cibulskis/refs/heads/main/data/AHRF2024.csv")|> 
  clean_names()

AHRF2025 <- read_csv("https://raw.githubusercontent.com/djnf-data-2026/laine-cibulskis/refs/heads/main/data/AHRF2025.csv")|> 
  clean_names()
```

## Starting to pull documentation for variables

```{r}
AHRF2024documentation <- read_excel("C:/Users/laine/OneDrive/Desktop/Code/DJNF/laine-cibulskis/documentation/AHRF 2023-2024 User Tech/AHRF 2023-2024 User Tech/AHRF 2023-2024 Technical DocumentationCSV_Feb2025.xlsx")
# Expenditures
AHRF2024documentation[c(3309, 3310, 3311, 3312, 3315, 3316, 3317, 3318), ]

```

## 2024 AHRF data: short term non-general and long term hospital expenditures by county
##### The AHRF 2024-2025 User Guide explains that short term non-general hospitals provide specialized care, and the majority of their patients stay for fewer than 30 days. Long term hospitals may provide either non specialized or specialized care, and the majority of their patients stay for 30 or more days.

```{r}
AHRF2024_MI |>
  select(cnty_name,stnglth_reprtd_fac_exp_thou_22) |>
  group_by(cnty_name) |>
  arrange(desc(stnglth_reprtd_fac_exp_thou_22)
)
```

### Per capita

```{r}
AHRF2024_MI |>
  select(cnty_name,stnglth_reprtd_fac_exp_thou_22, popn_est_22) |>
  group_by(cnty_name) |>
  summarise(pc_exp = ((stnglth_reprtd_fac_exp_thou_22 * 1000)/ popn_est_22)) |>
  arrange(desc(pc_exp)
)
```

### Per hospital in the county

```{r}
stnglth_exp_ph_22_MIc <- AHRF2024_MI |>
  select(cnty_name,stnglth_reprtd_fac_exp_thou_22, stnglth_rptng_exp_22) |>
  group_by(cnty_name) |>
  summarise(stnglth_exp_ph_22_MIc = ((stnglth_reprtd_fac_exp_thou_22 * 1000) / stnglth_rptng_exp_22)) |>
  arrange(desc(stnglth_exp_ph_22_MIc)
)
```

## 2024 AHRF data: short term general hospitals
##### The The AHRF 2024-2025 User Guide explains that short term general hospitals provide non specialized care, and the majority of their patients stay for fewer than 30 days.

```{r}
AHRF2024_MI |>
  select(cnty_name,stgh_reprtd_fac_exp_thou_22) |>
  group_by(cnty_name) |>
  arrange(desc(stgh_reprtd_fac_exp_thou_22)
)
```

### Per capita

```{r}
AHRF2024_MI |>
  select(cnty_name,stgh_reprtd_fac_exp_thou_22, popn_est_22) |>
  group_by(cnty_name) |>
  summarise(pc_exp = ((stgh_reprtd_fac_exp_thou_22 * 1000) / popn_est_22)) |>
  arrange(desc(pc_exp)
)
```

### Per hospital in the county

```{r}
stgh_exp_ph_22_MIc <- AHRF2024_MI |>
  select(cnty_name,stgh_reprtd_fac_exp_thou_22, stgh_rptng_exp_22
) |>
  group_by(cnty_name) |>
  summarise(stgh_exp_ph_22_MIc = ((stgh_reprtd_fac_exp_thou_22 * 1000) / stgh_rptng_exp_22
)) |>
  arrange(desc(stgh_exp_ph_22_MIc)
)
```

## 2025 AHRF data: short term non-general and long term hospital expenditures by county

```{r}
AHRF2025_MI |>
  select(cnty_name,stnglth_reprtd_fac_exp_thou_22) |>
  group_by(cnty_name) |>
  arrange(desc(stnglth_reprtd_fac_exp_thou_22)
)
```

### Per capita

```{r}
AHRF2025_MI |>
  select(cnty_name,stnglth_reprtd_fac_exp_thou_23, popn_est_23) |>
  group_by(cnty_name) |>
  summarise(pc_exp = ((stnglth_reprtd_fac_exp_thou_23 * 1000)/ popn_est_23)) |>
  arrange(desc(pc_exp)
)
```

### Per hosptial in the county

```{r}
AHRF2025_MI |>
  select(cnty_name,stnglth_reprtd_fac_exp_thou_23, stgh_rptng_exp_23) |>
  group_by(cnty_name) |>
  summarise(ph_exp = ((stnglth_reprtd_fac_exp_thou_23 * 1000) / stgh_rptng_exp_23)) |>
  arrange(desc(ph_exp)
)
```

## 2025 AHRF data: short term general hospitals

```{r}
AHRF2025_MI |>
  select(cnty_name,stgh_reprtd_fac_exp_thou_23) |>
  group_by(cnty_name) |>
  arrange(desc(stgh_reprtd_fac_exp_thou_23)
)
```

## Per capita

```{r}
AHRF2025_MI |>
  select(cnty_name,stgh_reprtd_fac_exp_thou_23, popn_est_23) |>
  group_by(cnty_name) |>
  summarise(pc_exp = ((stgh_reprtd_fac_exp_thou_23 * 1000) / popn_est_23)) |>
  arrange(desc(pc_exp)
)
```

## Per hospital in the county

```{r}
AHRF2025_MI |>
  select(cnty_name,stgh_reprtd_fac_exp_thou_23, stgh_rptng_exp_23) |>
  group_by(cnty_name) |>
  summarise(ph_exp = ((stgh_reprtd_fac_exp_thou_23 * 1000) / stgh_rptng_exp_23)) |>
  arrange(desc(ph_exp)
)
```

## Quick IL comparison

### Per hospital in each county

```{r}
AHRF2025_IL |>
  select(cnty_name,stgh_reprtd_fac_exp_thou_23, stgh_rptng_exp_23) |>
  group_by(cnty_name) |>
  summarise(ph_exp = ((stgh_reprtd_fac_exp_thou_23 * 1000) / stgh_rptng_exp_23)) |>
  arrange(desc(ph_exp)
)
```

### Washtenaw expenditure change math

```{r}
1014027286 - 984818286
```

# 6/1 data work: state and nation level

## 2025 AHRF: short term general hospital analysis continued

### Total short term general hospital expenditures by state (state nation data set does not have any expenditure data)

```{r}
AHRF2025 |>
  select(st_name_abbrev, stgh_reprtd_fac_exp_thou_23) |>
  group_by(st_name_abbrev) |>
  summarise(stgh_23_exp_by_state = (sum((stgh_reprtd_fac_exp_thou_23 * 1000)))) |>
  arrange(desc(stgh_23_exp_by_state)
  )
```

### Save the by state expenditures to use in other operations
```{r}
stgh_23_exp_by_state <- 
  AHRF2025 |>
  select(st_name_abbrev, stgh_reprtd_fac_exp_thou_23) |>
  group_by(st_name_abbrev) |>
  summarise(stgh_23_by_state = (sum((stgh_reprtd_fac_exp_thou_23 * 1000)))
  )
```

### View the short term general hosptials by state

```{r}
AHRF2025 |>
  select(st_name_abbrev, stgh_rptng_exp_23) |>
  group_by(st_name_abbrev) |>
  summarise(stgh_23_by_state = (sum(stgh_rptng_exp_23))) |>
  arrange(desc(stgh_23_by_state))
```

### Save the short term general hospitals by state to use in other operations

```{r}
stgh_23_by_state <- 
  AHRF2025 |>
  select(st_name_abbrev, stgh_rptng_exp_23) |>
  group_by(st_name_abbrev) |>
  summarise(stgh_23_by_state = (sum(stgh_rptng_exp_23))) |>
  arrange(desc(stgh_23_by_state)
  )
```

### Joining the two new data frames

```{r}
stgh_23 <-
merge(x = stgh_23_exp_by_state, y = stgh_23_by_state, by = "st_name_abbrev", all = TRUE)
```

### Total short term general hospital expenditures by state, per hospital in the state (don't need to run again because data frame was saved with new headers)

```{r}
#stgh_23 |>
  #mutate(stgh_23_exp_by_state_ph = (stgh_23_by_state.x / stgh_23_by_state.y)) |>
  #arrange(desc(stgh_23_exp_by_state_ph))
  #)
```

### Save total short term general hospital expenditures by state, per hospital in the state to use in other operations

```{r}

stgh_23 <-
  stgh_23 |>
    rename(
    stgh_23_exp_by_state = stgh_23_by_state.x,
    stgh_23_by_state = stgh_23_by_state.y
    )

stgh_23_exp_by_state_ph <-
  stgh_23 |>
  select(stgh_23_exp_by_state, stgh_23_by_state) |>
  mutate(stgh_23_exp_by_state_ph = stgh_23_exp_by_state / stgh_23_by_state) |>
  arrange(desc(stgh_23_exp_by_state_ph))
```

### Now stgh_23_exp_by_state_ph has all the variables. Save that into the stgh_23 data frame

```{r}
stgh_23 <- stgh_23_exp_by_state_ph
```

## Create national expenditure comparison

### General summary

```{r}
summary(stgh_23$stgh_23_exp_by_state_ph)
```

### Michigan is ranked No. 14 in the nation, excluding CT (to be addressed later), for per hospital short term general hospital expenditures by state.

### $342,074,250 is Michigan's per hospital value--exactly the third quartile. Meaning 3/4 of the U.S. states (again, minus CT), have lower per hospital expenditures for short term general hospitals.

### Michigan difference from mean calculation

```{r}
342074250 - 291256410
```

# 6/2 data work

## Join Michigan data by county for map

```{r}
full_join(
  x,
  y,
  by = NULL,
  copy = FALSE,
  suffix = c(".x", ".y"),
  ...,
  keep = NULL,
  na_matches = c("na", "never"),
  multiple = "all",
  relationship = NULL
)
```


## Short term non general and long term hospitals

## 2025 AHRF data: short term general hospitals

```{r}
AHRF2025_MI |>
  select(cnty_name,stgh_reprtd_fac_exp_thou_23) |>
  group_by(cnty_name) |>
  arrange(desc(stgh_reprtd_fac_exp_thou_23)
)
```

## Per capita

```{r}
AHRF2025_MI |>
  select(cnty_name,stgh_reprtd_fac_exp_thou_23, popn_est_23) |>
  group_by(cnty_name) |>
  summarise(pc_exp = ((stgh_reprtd_fac_exp_thou_23 * 1000) / popn_est_23)) |>
  arrange(desc(pc_exp)
)
```

## Per hospital in the county

```{r}
AHRF2025_MI |>
  select(cnty_name,stgh_reprtd_fac_exp_thou_23, stgh_rptng_exp_23) |>
  group_by(cnty_name) |>
  summarise(ph_exp = ((stgh_reprtd_fac_exp_thou_23 * 1000) / stgh_rptng_exp_23)) |>
  arrange(desc(ph_exp)
)
```

## Quick IL comparison

### Per hospital in each county

```{r}
AHRF2025_IL |>
  select(cnty_name,stgh_reprtd_fac_exp_thou_23, stgh_rptng_exp_23) |>
  group_by(cnty_name) |>
  summarise(ph_exp = ((stgh_reprtd_fac_exp_thou_23 * 1000) / stgh_rptng_exp_23)) |>
  arrange(desc(ph_exp)
)
```

### Washtenaw expenditure change math

```{r}
1014027286 - 984818286
```

## 2025 AHRF: short term non general hospitals and long term hospital analysis continued

### Total short term non general hospitals and long term hospital expenditures by state (state nation data set does not have any expenditure data)

```{r}
AHRF2025 |>
  select(st_name_abbrev, stnglth_reprtd_fac_exp_thou_23) |>
  group_by(st_name_abbrev) |>
  summarise(stnglth_23_exp_by_state = (sum((stnglth_reprtd_fac_exp_thou_23 * 1000)))) |>
  arrange(desc(stnglth_23_exp_by_state)
  )
```

### Save the by state expenditures to use in other operations
```{r}
stnglth_23_exp_by_state <- 
  AHRF2025 |>
  select(st_name_abbrev, stnglth_reprtd_fac_exp_thou_23) |>
  group_by(st_name_abbrev) |>
  summarise(stnglth_23_by_state = (sum((stnglth_reprtd_fac_exp_thou_23 * 1000)))
  )
```

### View the short term general hosptials by state

```{r}
AHRF2025 |>
  select(st_name_abbrev, stnglth_rptng_exp_23) |>
  group_by(st_name_abbrev) |>
  summarise(stgh_23_by_state = (sum(stnglth_rptng_exp_23))) |>
  arrange(desc(stnglth_23_by_state))
```

### Save the short term general hospitals by state to use in other operations

```{r}
stnglth_23_by_state <- 
  AHRF2025 |>
  select(st_name_abbrev, stnglth_rptng_exp_23) |>
  group_by(st_name_abbrev) |>
  summarise(stgh_23_by_state = (sum(stnglth_rptng_exp_23))) |>
  arrange(desc(stgh_23_by_state)
  )
```

### Joining the two new data frames

```{r}
stnglth_23 <-
merge(x = stgh_23_exp_by_state, y = stgh_23_by_state, by = "st_name_abbrev", all = TRUE)
```

### Total short term general hospital expenditures by state, per hospital in the state

```{r}
stnglth_23 |>
  mutate(stgh_23_exp_by_state_ph = (stgh_23_by_state.x / stgh_23_by_state.y)) |>
  arrange(desc(stgh_23_exp_by_state_ph)
  )
```

### Save total short term general hospital expenditures by state, per hospital in the state to use in other operations

```{r}

stnglth_23 <-
  stgh_23 |>
    rename(
    stgh_23_exp_by_state = stgh_23_by_state.x,
    stgh_23_by_state = stgh_23_by_state.y
    )

stgh_23_exp_by_state_ph <-
  stgh_23 |>
  select(stgh_23_exp_by_state, stgh_23_by_state) |>
  mutate(stgh_23_exp_by_state_ph = stgh_23_exp_by_state / stgh_23_by_state) |>
  arrange(desc(stgh_23_exp_by_state_ph))
```

### Now stgh_23_exp_by_state_ph has all the variables. Save that into the stgh_23 data frame

```{r}
stgh_23 <- stgh_23_exp_by_state_ph
```

## Create national expenditure comparison

### General summary

```{r}
summary(stgh_23$stgh_23_exp_by_state_ph)
```

## Datawrapper



### Load libraries and API key

```{r}
library(DatawRappr)
library(tidyverse)
datawrapper_auth(api_key =  "qWGtNy2YSjg6j2gfzgCSQ3iYFeeciWYTJTuWyVfTfWqWI7jbNHs9HOcmYIOsc6X2", overwrite=TRUE)
```

### Call in data and display head

my_chart <- dw_create_chart()

  rename(
    State = st_name_abbrev
    Revenue, short term general hospitals = stgh_23_by_state.x
    Total short term general hospitals = stgh_23_by_state.y